---
title: "perceptron"
author: "Joyce Robbins"
date: "8/8/2018"
output: 
  html_document:
    code_folding: hide
    fig_height: 4
    fig_width: 6
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(perceptron)
knitr::opts_chunk$set(echo = TRUE)

#install_github("jtr13/perceptron")
```

The perceptron is a simple algorithm that learns to classify inputs into two classes by adjusting the weights (w) in the equation $y_i =$ sign$(w_ix_i)$ until all inputs in a training set are correctly classified. Here the steps of algorithm will be presented visually in two-dimensional space.

### The setup

We start by plotting $(x_1, x_2)$, coloring each point by class. Note that the points can be separated by a line; if this is not the case, the algorithm won't work.

```{r}
X <- matrix(c(-1, 1, 2, 4, 6, 4, 6, 6, 5, 8,
                -3, 7, 4, 4, 2, 8, 6, 8, 10, 10)/5, 10)

Y <- c(-1, -1, -1, -1, -1, 1, 1, 1, 1, 1)
X <- cbind(rep(1, length(Y)), X)
draw_points(X[,2], X[,3], Y, axes = TRUE)
```


We start with an arbitrary weight vector, $(w0, w1, w2)$.  Often $(0, 0, 0)$ is used, but we'll start with $(0, .5, .5)$ so we can visualize it:

```{r weightvector}
W <- c(0, .5, .5)
s <- shift(W)
draw_points(X[,2], X[,3], Y, axes = TRUE)
mtext("X1", side = 1, line = 3)
mtext("X2", side = 2, line = 3)
draw_weight_vector(W)
label_vector(s["x"], s["y"], W[2] + s["x"],
             W[3] + s["y"], "weight")
```

Next we circle any point for which $y_i \neq$ sign$(w_ix_i)$ -- these are the misclassified points.

```{r circlepoints}
<<weightvector>>
mis.points <- X[sign(W %*% t(X)) != Y, , drop = F]
points(mis.points[,2], mis.points[,3], 
       col = "blue", cex = 2)
```


Note that visually, *the line orthogonal to the weight vector* is the decision boundary between the two classes.  This is true because for points on the line, sign $(w_ix_i) = 0.$

```{r}
<<circlepoints>>
draw_hyperplane(W)
e <- get_endpoints(W)
label_vector(e[1], e[2], e[3], e[4],
             label = "hyperplane",
             "below")
```

